---
title: "Dundreggan2025"
author: "Petra Guy"
date: "`r Sys.Date()`"
output: html_document
---


```{r, message=T, warning=F}
library(readxl)
library(dplyr)
library(tidyr)
library(ggpubr)
library(ggplot2)

setwd('C:/dev/code/Petra')
knitr::opts_chunk$set(
  echo = F,      # hide all code by default
  message = FALSE,   # hide package messages
  warning = FALSE    # hide warnings
)
```

Dundreggan has been measured at 4 time points, August 2023, April 2024, November 2024 and October 2025
Initially tree tags were lost and there was no tree_id between years. Different numbers of trees 
were recorded each year. In April 2024 we made some attempt to tag trees with numbered tags start to follow exact tree growth rates. 

This first quick exploration just uses plot averages - that is - we ignore growth rates between trees and only look at change in average plot height of all trees. 

```{r}
#read in the data
# Path to your Excel file of just height data over all 4 time points
file_path <- "../../data/field/Dundreggan2025.xlsx"

# Get all sheet names - eahc sheet is a different plot, site 1 fert, site 1 pellet etc
sheet_names <- excel_sheets(file_path)

# Read all sheets into a named list of data frames
dfs <- lapply(sheet_names, function(sheet) read_excel(file_path, sheet = sheet))
names(dfs) <- sheet_names

```


