---
title: "TilhillOlivia2025"
author: "Petra Guy"
date: "`r Sys.Date()`"
output: pdf_document
---
```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(readxl)
  library(dplyr)
  library(tidyr)
  #library(ggpubr)
  library(ggplot2)
  #library(nlme)
  library(kableExtra)
  library(stringr)
  library(ggpattern) 
  library(patchwork)
  #library(lme4)
  #library(lmerTest)
 # library(broom.mixed)
  #library(emmeans)
  library(purrr)
  #library(car)
  #library(MASS)
})


knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```

```{r, echo=FALSE, warning=FALSE, fig.height=4, fig.width=7, fig.pos='H'}
#read raw data
data = read_excel('../../../data/field/TilhillOlivia2025.xlsx')

sheets <- excel_sheets('../../../data/field/TilhillOlivia2025.xlsx')  # get sheet names

# read all sheets into a named list, suppressing the New names messages
all_sheets <- suppressMessages(
  lapply(sheets, function(s) {
    read_excel('../../../data/field/TilhillOlivia2025.xlsx', sheet = s)
  })
)
names(all_sheets) <- sheets

# remove meta sheets
all_sheets <- all_sheets[-c(1,2)]

#fix the colnames issue
all_sheets <- map(
  all_sheets,
  ~ .x %>%
    rename_with(~ str_replace(.x, "Height\\s+_cm_", "Height_cm_"))
)

#add the delta_h in now
all_sheets <- map(all_sheets, ~ .x %>%
                    mutate(
                      delta_h   = Height_cm_1 - Height_cm_0,
                      delta_rcd = Rcd_mm_1 - Rcd_mm_0
                    ))
                   

selected_cols <- all_sheets_long <- map(all_sheets, ~ .x %>%
                         dplyr::select(Tree_Species, Comment_1, Treatment, Health_1, delta_h, delta_rcd)) 


#make long df
df_long <- bind_rows(
  selected_cols,
  .id = "Plot"
)
r
#change teatment names
df_long <- df_long %>%
  mutate(
    Treatment = case_when(
      Treatment == "Fungi_Control" ~ "Control",
      Treatment == "Treatment"     ~ "Pellet",
      TRUE                         ~ Treatment  # keep all other values
    )
  )

df_long_noDt <- df_long %>%
  filter(
    is.na(Comment_1) |
    !str_detect(
      Comment_1,
      regex("\\bDt\\b|\\bdead\\s+top\\b|\\bDying\\b", ignore_case = TRUE)
    )
  )

```

INTERNAL DOCUMENT ONLY

SUMMARY

INTRODUCTION

This Tillhill site is MBR planting where mix of trees spp spread over multiple plots. Not a balanced design for analysis, but we will be able to pull out enough from some plots to use plot as a random effect. Might have fewer random than ideal - i.e. < 5. 

TREATEMENTS
Control/Treatment across 9 plots - paired with MBR in each
Lac am. 

METHODS
Block 11. Larger -ves are health 0. ~ 6 trees -5 < delta_h < 0. Prob omit these and leave the 0
Block 12 - many larger -ve change in ht due to Dt - remove all rows with Comment_1 contains dead top or Dt, many of these are health_0 anyway, a few are not, changed one pine with health 2 and height_1 = 0
Block 13. Some large -ves with health 2, some comment as dead top - remove rows with dead top or Dt

Some -ves are because if ht is entered as 0 or NA , when 0, delta h is calculated, if NA, then its not

ANALYSIS
```{r, echo=FALSE, warning=FALSE, fig.height=7, fig.width=7, fig.pos='H'}
#explore data - what do -ve heights look like etc and how many trees per plot etc.

p1 = ggplot(df_long, aes(x = factor(Health_1), y = delta_h)) +
  geom_boxplot(fill = "skyblue", color = "black") +  # color & fill
   geom_hline(yintercept = 0, linetype = "dashed", color = "red") +  # horizontal line at 0
  labs(
    x = "Health class",
    y = "Change in height (cm)",
   ) +
  theme_bw() +
  theme(
    text = element_text(size = 12)
  )

#this is after the Dt and dead top have been removed
p2 = ggplot(df_long_noDt, aes(x = factor(Health_1), y = delta_h)) +
  geom_boxplot(fill = "skyblue", color = "black") +  # color & fill
   geom_hline(yintercept = 0, linetype = "dashed", color = "red") +  # horizontal line at 0
  labs(
    x = "Health class",
    y = "Change in height (cm)",
   ) +
  theme_bw() +
  theme(
    text = element_text(size = 12)
  )

 
p1 / p2
```

Most negative change in height in health_0, as expected - three were large positive delta h in health 0 - these must have been errors, and health changed to 2. There was also 1 pine with health NA, but heights recorded, this must be error and changed to health_2. The remaining Health NA are lost trees. For remaining trees some are dead top, all these need to be removed as the change in height cannot be assessed. 

```{r, echo=FALSE, warning=FALSE, fig.height=7, fig.width=7, fig.pos='H'}
#explore negative some more
#leave out the health 0 as we know they are all -ve
df_neg <- df_long_noDt %>%
  filter( Health_1 >0,
         delta_h <= 0)

# Quick histogram of negative delta_h values
p1 = ggplot(df_neg, aes(x = delta_h)) +
  geom_histogram(binwidth = 1, fill = "firebrick", color = "black") +
  scale_x_continuous(breaks = seq(floor(min(df_neg$delta_h)),
                                  ceiling(max(df_neg$delta_h)), by = 1)) +
  labs(title = "Distribution of negative delta_h values for all tree spp",
       subtitle = "Health_1 =1 or 2",
       x = "delta_h (cm) (negative values only)",
       y = "Count") +
  theme_minimal()
#overlapping bars for control and treatment
ggplot(df_neg, aes(x = delta_h, fill = Treatment)) +
  geom_histogram(binwidth = 1, color = "black",
                 position = position_dodge(width = 1), alpha = 0.8) +
  scale_x_continuous(breaks = seq(floor(min(df_neg$delta_h)),
                                  ceiling(max(df_neg$delta_h)), by = 1)) +
  scale_fill_manual(values = c("Control" = "orange", "Pellet" = "forestgreen")) +
  labs(title = "Distribution of negative delta height values for all tree spp",
       subtitle = "Health_1 = 1 or 2",
       x = "change in height (cm) (negative values only)",
       y = "Count",
       fill = "Treatment") +
  theme_minimal()

#cumulative totals

# Define bins in the order you want
bins <- 0:-5  # 0, -1, -2, ..., -5

# Compute counts and cumulative counts per Treatment
df_cum <- df_neg %>%
  group_by(Treatment) %>%
  summarise(count = table(factor(delta_h, levels = bins))) %>%
  tidyr::unnest(cols = c(count)) %>%
  group_by(Treatment) %>%
  mutate(
    bin = bins,
    cum_count = cumsum(count)
  ) %>%
  ungroup()

p2 = ggplot(df_cum, aes(x = bin, y = cum_count, color = Treatment)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = bins) +
  scale_color_manual(values = c("Control" = "orange", "Pellet" = "forestgreen")) +
  labs(
    title = "Cumulative frequency of negative delta h for all tree spp",
    subtitle = "Health_1 = 1 or 2",
    x = "change in height (cm) threshold",
    y = "Cumulative count",
    color = "Treatment"
  ) +
  theme_minimal()



p1 /p2

```

Frequencies of counts of negative delta_h
The large negative values will be removed anyway - but looking at the smaller ones, -1 and 0, where we know our measurement error lies. If we leave out anything 0 and below, we lose 7 treatment and 4 control trees. If we include -1, we lose 10 treatment and 5 control trees. SInce there arent many trees at this site in each block - look at block summaries if we keep 0s, -1s, etc.

The tables show the number of control and treatment trees in each block if we retain i) delta h >= 0, or secondly - delta_h > -1. Increasing the range of delta h to -1 adds 5 trees, spread across the different plots. By increasing the range it doesnt mean we then have a tree sp or block which can incorporate into the analysis which wasnt there if delta_h >= 0. 

Therefore we will carry on with data filtered s.t health_1 >=1 and delta_h >= 0 

```{r, echo=FALSE, warning=FALSE, fig.height=9, fig.width=7, fig.pos='H'}
# Summarize number of trees per Tree_species × Plot

#this for delta h includes zeros
tree_counts0 <- df_long_noDt %>%
  filter(Health_1 > 0, delta_h >= 0) %>%
  group_by(Tree_Species, Plot, Treatment) %>%
  summarise(n_trees = n(), .groups = "drop") %>%
  arrange(Tree_Species, Plot)

# Pivot to wide format: plots as columns
tree_counts_wide0 <- tree_counts0 %>%
  pivot_wider(names_from = Plot, values_from = n_trees, values_fill = 0)


tree_counts_wide0
 #this for delta h strictly > 0
tree_counts1 <- df_long_noDt %>%
  filter(Health_1 >0, delta_h >=-1) %>%
  group_by(Tree_Species, Plot, Treatment) %>%
  summarise(n_trees = n(), .groups = "drop") %>%
  arrange(Tree_Species, Plot)

# Pivot to wide format: plots as columns
tree_counts_wide1 <- tree_counts1 %>%
  pivot_wider(names_from = Plot, values_from = n_trees, values_fill = 0)


tree_counts_wide1
```

If we filter health >=1 and delta h >= 0 this leaves multiple plots per tree species, but not 3.

Alnus : 2 plots \newline
Betula: 3 plots \newline
Corylus: 2 plots \newline
Pinus: 2 plots \newline
Quercus 4 plots \newline

Therefore we will first change in heights for alnus and corylus. If the data looks interesting, maybe test, even though set up not ideal. \newline
For Betula and oak we model with plot as random \newline

Start with a model for Betula.

```{r, echo=FALSE, warning=FALSE, fig.height=7, fig.width=7, fig.pos='H'}

#box plots and bar plots of change in height - includes the zeros

#get the birch data
#filter for health > 0 and delta_h >=0
BirchPlots = c('Block 11', 'Block 12', 'Block 14')
birch <- df_long_noDt %>%
  filter(
    Tree_Species == "Betula_pubescens",
    Health_1 > 0,
    delta_h >= 0,
    Plot %in% BirchPlots
  )



p1 = ggplot(birch, aes(x = Treatment, y = delta_h, fill = Treatment)) +
  geom_boxplot(width = 0.7, outlier.shape = 21, color = "black") +
  facet_wrap(~ Plot, scales = "free_y") +  # one facet per species
  scale_fill_manual(values = c(Control = "orange", Pellet = "forestgreen" )) +
  labs(
    title = "Distribution of delta_h by Treatment and Plot, Betula",
    x = "Treatment",
    y = "Change in Height (cm)"
  ) +
  theme_bw() +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "none"
  )

# summary data for the positive only

df_summary_delta_h <- birch %>%
  group_by(Plot, Treatment) %>%
  summarise(
    mean_delta_h = mean(delta_h, na.rm = TRUE),
    se_delta_h   = sd(delta_h, na.rm = TRUE) / sqrt(n()),
    n            = n(),
    .groups = "drop"
  )

p2 = ggplot(df_summary_delta_h, aes(x = Treatment, y = mean_delta_h, fill = Treatment)) +
  geom_col(width = 0.7) +
  geom_errorbar(aes(ymin = mean_delta_h - se_delta_h, ymax = mean_delta_h + se_delta_h),
                width = 0.3, color = "black") +
  facet_wrap(~ Plot, scales = "free_y") +
  scale_fill_manual(values = c(
    Control = "orange",
    Pellet  = "forestgreen"
  )) +
  labs(
    title = "Mean delta_h  SE by Treatment and Plot, Betula",
    x = "Treatment",
    y = "Mean Change in Height (cm)"
  ) +
  theme_bw() +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "none"
  )
p1/p2
```

```{r, echo=FALSE, warning=FALSE,message = FALSE, results = 'hide',fig.height=4, fig.width=7, fig.pos='H'}
#Bayesian model


BirchPlots = c('Block 11', 'Block 12', 'Block 14')
birch <- df_long_noDt %>%
  filter(
    Tree_Species == "Betula_pubescens",
    Health_1 > 0,
    delta_h >= 0,
    Plot %in% BirchPlots
  )
df_model =  birch %>% mutate(delta_h_pos = delta_h + 0.1)

#+0.1 to get rid of zeros for lognormal


suppressMessages({
  library(brms)
  library(tibble)
})



#posterior means and credibilty intervals
m_gauss <- brm(
  delta_h ~ Treatment + (1 | Plot),  # random intercept for Plot
  data = birch,
  family = gaussian(),
  chains = 4, cores = 4
)

#this extracts the effect size for the pellet (i.e, intercept+pellet effect)
#since this is lognormal, actual effect size is exp(intercept+pellet)
summary = posterior_summary(m_gauss, probs = c(0.025, 0.975))

#need to back transform because used a log-normal
#b_intercept is the basline or control, b_Treatment is difference in log(delta_h) due to pellet



#  Extract posterior samples- this gives a df of effect sizes based on the distribution
#for the effect sizes from the model
post <- posterior_samples(m_gauss)
n <- nrow(post)

#get the control (intercept) effect, and transform back from lognormal
control_vals <- exp(post$b_Intercept) 
median_control <- median(control_vals)
ci_control <- quantile(control_vals, probs = c(0.025, 0.975))

pellet_vals <- exp(post$b_Intercept + post$b_TreatmentPellet)
median_pellet <- median(pellet_vals)
ci_pellet <- quantile(pellet_vals, probs = c(0.025, 0.975))

pct_change <- (pellet_vals / control_vals - 1) * 100
median_pct <- median(pct_change)
ci_pct <- quantile(pct_change, c(0.025, 0.975))



summary_table <- tibble(
  Parameter = c("Control (Intercept)", "Pellet", "Pellet % change"),
  Median = c(median_control, median_pellet, median_pct),
  CI_lower = c(ci_control[1], ci_pellet[1], ci_pct[1]),
  CI_upper = c(ci_control[2], ci_pellet[2], ci_pct[2])
)

print(summary_table)

#lets get a probability of positive change
#this gives the number of TRUE for mean of treatment effect is > 0
prob_positive = mean(post$b_TreatmentPellet >0)

###############################################################################
#MODEL VALIDATION - JUST LOOKING AT RESIDUALS VERUS ACTUAL AND PLOT OF ACTUAL MEAN VERSUS
#THE MODELLED MEANS

#this looks at fitting method - need rhat = 1
#print(m)


#rediduals versus fitted
# Posterior predicted values (median for each observation)
y_pred <- apply(posterior_predict(m_gauss), 2, median)

# Observed values
y_obs <- df_model$delta_h_pos

# Residuals
resid <- y_obs - y_pred

# Basic residuals vs predicted plot
plot(y_pred, resid,
     xlab = "Predicted change in height (median posterior)",
     ylab = "Residuals",
     main = "Residuals vs Predicted for Quercus Model",
     pch = 16, col = "steelblue")
abline(h = 0, col = "red", lty = 2)

#look at model predictions for mean of control and mean of treatment
# Simulate posterior predictions, from each start vlaue of delta he, predict
#on the model geneates a set of new delta ha's
y_rep <- posterior_predict(m_gauss)  # matrix: draws x observations

# Indices for each treatment
ctrl_idx <- which(df_model$Treatment == "Control")
pellet_idx <- which(df_model$Treatment == "Pellet")

# Compute mean per draw for each group of simulations
mean_ctrl <- apply(y_rep[, ctrl_idx], 1, mean)
mean_pellet <- apply(y_rep[, pellet_idx], 1, mean)

# Observed means
obs_mean_ctrl <- mean(df_model$delta_h_pos[ctrl_idx])
obs_mean_pellet <- mean(df_model$delta_h_pos[pellet_idx])

# Plot
hist(mean_ctrl, breaks = 30, col = rgb(0,0,1,0.3), main="Posterior predictive means", xlab=" change in height")
abline(v = obs_mean_ctrl, col = "blue", lwd = 2)
hist(mean_pellet, breaks = 30, col = rgb(1,0,0,0.3), add = TRUE)
abline(v = obs_mean_pellet, col = "red", lwd = 2)
legend("topright", legend=c("Control","Pellet"), col=c("blue","red"), lwd=2)

```
